extinction risk makefile, compiliation notes

### 2019-10-07 MainText.Rmd
some comments starting the process of gettin this paper ready for Ecosphere

@@

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annal survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--Earlier study on metapopulation extinction risk : Large-scale spatial synchrony may also be generated by
other processes such as migration and predation, but regional
stochasticity in the form of spatially correlated weather conditions is
probably the dominant synchronizing mechanism (ref: Nature1998) but what about taildep. effect on extrisk?.-->
In the metapopulation context, regional stochasticity 
can simultaneously affected multiple habitat patches where spatially correlated weather conditions were 
probably the dominant synchronizing mechanism for extinctions [@hanski1998metapopulation]. A 21 years-long
study by @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased amplitude and span of spatial synchrony 
over a metapopulation network of 4000 dry meadows and could affect the long-term population viability.
Though earlier literature considered the effect of spatial and/ or temporal correlation in estimating extinction eisk of a metapopulation, but to our knowledge, using the concept of "tail dependence" (i.e. correlation would be stronger during extreme years) to model extinction risk remains unexplored.

@@

<!--brief intro for copula: I am not sure whether it is needed in this paper or not?-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families with asymmetric tail that are used to model exclusively 
left and right tail dependence between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of dependence 
depending on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail dependence have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

>>>>

<!--***DAN: 
1) Shyamolina,  was reminded during my editing of BIVAN that "tail dependence" has 
a specific mathematical definition (now given in the Background section of BIVAN), and that
is not what we mean here when we say "tail dependence". So we cannot go on calling it "tail
dependence" in this manuscript - that would be mike me calling my house "the white house" - 
there already is one of those, and it is much better known than my house, so we cannot
attempt to hijack the name. So can you change terminology throughout, please? Let's refer to
it using terminology such as "associations in the tails" or 
"tail associations". I would like you to provide a 
definition at an appropriate place early in the manuscript that we will call a pair
of positively associated variables "left-tail associated" if their association in their 
left tails is stronger than their association in their right tails (in a sense to be defined
mathematically in Methods, and then in methods we will define things based on cor_l-cor_u).
We call a pair of positively associated variable "right-tail associated" if their association
in their right tails is stronger than their association in their left tails.

2) You need to start getting this paper ready for submission - if you want to send it to 
the Ecosphere special issue (what do you think of that?) then we have to get moving. Please 
let me know soon, after reading about Ecospere on their website. If you think this is a 
good idea, let me know and I will send a summary of the paper to Jon Walter to make sure he is still willing to have us submit the paper to him, and to make sure the paper is close enough
to the theme of the special issue that he is going for, so that he will be willing to see it
submitted there.

3) Please
add the figure you have mentioned in the text but that does not exist. Please run a spell
check. Please go and check every sentence throughout the manuscript for grammatical 
errors. I think you will find some 
and will be able to correct them if you focus on that - there are many, and by correcting them
you will not only make the paper better you will get better at writing grammatically correct 
sentences in English.

4) Assuming you want to go for Ecosphere and Jon is OK with us submitting there, please read the "guidelines for authors" section of the Ecosphere website 
and format the paper for Ecosphere.

5) Please make all the improvements that you can see for yourself
are needed! I suspect if you read the paper carefully now after having worked on something
else for a while you will see many things you can improve!
-->

@@

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) 
<!--***DAN: Shyamolina, I do not know what the parenthesis refers to -->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annal survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction. 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

<!--Earlier study on metapopulation extinction risk : Large-scale spatial synchrony may also be generated by
other processes such as migration and predation, but regional
stochasticity in the form of spatially correlated weather conditions is
probably the dominant synchronizing mechanism (ref: Nature1998) but what about taildep. effect on extrisk?.-->
In the metapopulation context, regional stochasticity 
can simultaneously affected multiple habitat patches where spatially correlated weather conditions were 
probably the dominant synchronizing mechanism for extinctions [@hanski1998metapopulation]. A 21 years-long
study by @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased amplitude and span of spatial synchrony 
over a metapopulation network of 4000 dry meadows and could affect the long-term population viability.
Though earlier literature considered the effect of spatial and/ or temporal correlation in estimating extinction eisk of a metapopulation, but to our knowledge, using the concept of "tail dependence" (i.e. correlation would be stronger during extreme years) to model extinction risk remains unexplored.
<!--***DAN: Shyamolina, you cannot intro tail dependence (please change to calling it
tail association) in the last sentence of a paragraph. The reader does not know what that is, 
and a whole paragraph is needed to intro it.-->

<!--***DAN: Shyamolina, you need to make two major changes in how tail associations
are introdcued. 
  First, do it prior to saying anything about copulas. We have been wanting
to disassociate this paper, to some extent, from BIVAN. Well, probably the best way
to do that is by introducing tail dependence first, before copulas. You can 
describe it intuitively as association in the tails, and you can use pictures to describe it.
You can introduce the partial Spearman correlation (briefly in the Intro, more detail in Methods). 
You will need an explnation of why we 
use Spearman instead of Pearson (talk to me if you do not know, but you may want to read the 
new Background section of BIVAN again first). One does not need to know anything about copulas
to understand tail associations.
  Second, the introduction to copulas that is given is outmoded. See the BAckground section
to BIVAN for improvements. But you do not want to say anything like that much about copulas 
here, you just want to briefly link to the concept, cite some of the Intro work that
has been done (includeing BIVAN and some of the paper BIVAN cites for Intro work),
mention that tail associations are one aspect of copula structure that we have found important,
and introduce any specific things you will need, like the Clayton copula.

Right now this is quite a short paper. You have space to explain things and show pedagogical
figures.
-->

@@

<!--brief intro for copula: I am not sure whether it is needed in this paper or not?-->
<!--***DAN: Shyamolina, to answer your above Q, well, you use Clayton and survival 
Clayton in your simulations, so you probably have to explain copulas enough to explain those.-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families 
<!--***DAN: Shyamolina, we cannot just mention these, no one will know what they are.
As I said I think we should intro tail associations without copulas, and then subsequently link 
to copulas afterwards.-->
with asymmetric tail that are used to model exclusively 
left and right tail dependence between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of dependence 
depending on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail dependence have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

### 2019-10-07 makefile.Rmd
checkpoint issue solved
some comments starting the process of gettin this paper ready for Ecosphere

---
title: "makefile, compiliation notes"
author: "Daniel Reuman"
date: ""
output: pdf_document
---

@@

```{r tex_for_supp_mat}
library(rmarkdown)
rmarkdown::render(input="SuppMat.Rmd") 
```

>>>>

---
title: "extinction risk makefile, compiliation notes"
author: "Daniel Reuman"
date: ""
output: pdf_document
---

@@

```{r tex_for_supp_mat}
rmarkdown::render(input="SuppMat.Rmd") 
```
